---
title: "Bacterial Resetting of D/L Ratios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bacterial Resetting of D/L Ratios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(AARP)
library(ggplot2)
library(dplyr)
```

## Biological reworking of amino acids

Braun et al. (2017) showed that bacteria in marine sediments continuously
recycle amino acids: they consume D-form amino acids from necromass and
synthesize new L-form amino acids, pulling D/L ratios below the values that
pure abiotic racemization would predict. The magnitude of this effect scales
with organic carbon availability, which decreases with depth as organic matter
is degraded.

The AARP model now includes an optional bacterial resetting term. At each
integration timestep `dt`, the accumulated D/L in power-law space (Rx) evolves
as:

```
Rx(t+dt) = Rx(t)
           + dRacPL(kt, dt)                           # abiotic racemization
           - f_bac * (TOC(z) / TOC_surface) * Rx * dt  # bacterial resetting
```

The single free parameter `f_bac` (yr^-1) is the surface resetting rate.
Local TOC normalised to the surface value scales the effect with depth: high
near the surface where organic matter is abundant; low in old, organic-depleted
sediment. Based on necromass turnover times of hundreds to thousands of years
reported by Braun et al. (2017), plausible values range from 1e-4 to 1e-2
yr^-1.

## TOC profile

We build a synthetic but realistic downcore TOC profile using an exponential
decay model. TOC drops steeply in the top ~50 cm, then levels off to a
recalcitrant baseline:

```{r toc_profile}
# Exponential decay: TOC_0 = 5% at surface, asymptote ~0.3%
depths_toc <- c(0, 10, 25, 50, 100, 200, 350, 500)
toc_vals   <- 0.3 + 4.7 * exp(-depths_toc / 60)

toc_fn <- make_toc_model(depth_cm = depths_toc, toc_pct = toc_vals)

data.frame(depth_cm = seq(0, 500, 5),
           toc      = toc_fn(seq(0, 500, 5))) |>
  ggplot(aes(x = toc, y = depth_cm)) +
  geom_line(colour = "sienna", linewidth = 1) +
  scale_y_reverse() +
  labs(x = "TOC (%)", y = "Depth (cm)",
       title = "Synthetic TOC profile (exponential decay)") +
  theme_bw()
```

## Effect on D/L profiles

Compare D/L profiles for a range of `f_bac` values spanning the estimates from
Braun et al. (2017). All scenarios use the same temperature (5 deg C constant)
and sedimentation rate (50 cm/ka):

```{r dl_profiles}
am     <- rate_to_age_model(rate_cm_per_ka = 50, max_depth_cm = 500)
depths <- seq(0, 500, by = 10)

f_bac_vals <- c(0, 1e-4, 1e-3, 5e-3)
labels     <- c("f_bac = 0 (abiotic)", "f_bac = 1e-4", "f_bac = 1e-3", "f_bac = 5e-3")

res <- lapply(seq_along(f_bac_vals), function(j) {
  out <- racemize_depth_series(depths,
                               age_model = am,
                               temp_model = 5,
                               toc_model  = toc_fn,
                               f_bac      = f_bac_vals[j])
  out$scenario <- labels[j]
  out
})

do.call(rbind, res) |>
  mutate(scenario = factor(scenario, levels = labels)) |>
  ggplot(aes(x = DL, y = depth_cm, colour = scenario)) +
  geom_line(linewidth = 1) +
  scale_y_reverse() +
  scale_colour_brewer(palette = "Reds") +
  labs(x = "D/L", y = "Depth (cm)", colour = NULL,
       title = "Bacterial resetting suppresses D/L relative to abiotic model",
       subtitle = "T = 5 deg C constant, 50 cm/ka sedimentation rate") +
  theme_bw() +
  theme(legend.position = "bottom")
```

The strongest resetting occurs near the surface where TOC is highest. At depth,
as TOC approaches the recalcitrant baseline, the bacterial effect diminishes and
profiles converge toward the abiotic curve.

## Implied temperature bias

If bacterial reworking is real but ignored during temperature reconstruction,
the forward model will attribute lower-than-expected D/L to a cooler
temperature rather than to biological resetting. This creates an apparent
cooling bias. We can quantify this by asking: at what temperature would the
abiotic model produce the same D/L profile as a bacterially corrected model at
5 deg C?

```{r temp_bias}
# True scenario: 5 deg C + bacterial resetting at f_bac = 1e-3
true_with_bac <- racemize_depth_series(depths, am, temp_model = 5,
                                        toc_model = toc_fn, f_bac = 1e-3)

# Find apparent temperature (abiotic model) that minimises squared deviation
# at each depth independently, using a simple grid search
T_grid <- seq(1, 10, by = 0.1)
apparent_T <- sapply(seq_along(depths), function(di) {
  target_DL <- true_with_bac$DL[di]
  if (target_DL == 0) return(NA_real_)
  abiotic_DL <- sapply(T_grid, function(T) {
    racemize_depth_series(depths[di], am, temp_model = T)$DL
  })
  T_grid[which.min((abiotic_DL - target_DL)^2)]
})

bias_df <- data.frame(
  depth_cm   = depths,
  apparent_T = apparent_T,
  true_T     = 5,
  bias_C     = apparent_T - 5
) |> filter(!is.na(apparent_T))

ggplot(bias_df, aes(x = bias_C, y = depth_cm)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_line(colour = "steelblue", linewidth = 1) +
  scale_y_reverse() +
  labs(x = "Apparent temperature bias (deg C)", y = "Depth (cm)",
       title = "Apparent cooling bias from ignoring bacterial reworking",
       subtitle = "f_bac = 1e-3 yr^-1, T = 5 deg C true") +
  theme_bw()
```

The bias is largest in young, shallow sediment (where TOC is high and bacterial
activity is greatest) and diminishes with depth. For `f_bac = 1e-3` yr^-1, the
surface bias can reach 1-2 deg C, while old, deep sediment converges toward
the true temperature.

## Sensitivity to TOC decay rate

The shape of the TOC profile matters. A faster decay (TOC drops quickly to
background) concentrates the bacterial effect near the surface. A slower decay
(gradual decline to background) produces a more uniform correction:

```{r toc_sensitivity}
toc_fast <- make_toc_model(depths_toc, 0.3 + 4.7 * exp(-depths_toc / 20))
toc_slow <- make_toc_model(depths_toc, 0.3 + 4.7 * exp(-depths_toc / 150))

scens <- list(
  list(fn = toc_fast, label = "Fast decay (e-fold 20 cm)"),
  list(fn = toc_fn,   label = "Medium decay (e-fold 60 cm)"),
  list(fn = toc_slow, label = "Slow decay (e-fold 150 cm)")
)

res2 <- lapply(scens, function(s) {
  out <- racemize_depth_series(depths, am, temp_model = 5,
                               toc_model = s$fn, f_bac = 1e-3)
  out$scenario <- s$label
  out
})

abiotic <- racemize_depth_series(depths, am, temp_model = 5)
abiotic$scenario <- "Abiotic (no correction)"

bind_rows(do.call(rbind, res2), abiotic) |>
  mutate(scenario = factor(scenario,
    levels = c("Abiotic (no correction)", "Fast decay (e-fold 20 cm)",
               "Medium decay (e-fold 60 cm)", "Slow decay (e-fold 150 cm)"))) |>
  ggplot(aes(x = DL, y = depth_cm, colour = scenario)) +
  geom_line(linewidth = 1) +
  scale_y_reverse() +
  scale_colour_brewer(palette = "Set1") +
  labs(x = "D/L", y = "Depth (cm)", colour = NULL,
       title = "TOC decay rate controls depth of bacterial influence",
       subtitle = "f_bac = 1e-3 yr^-1 in all corrected scenarios") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Lake sediment example: humped TOC profile

Marine sediment TOC profiles almost always decrease monotonically with depth
as organic matter degrades. Lake sediments can behave very differently. A
lake whose productivity varied through the Holocene — perhaps peaking during
a warm, wet mid-Holocene optimum — can show a TOC profile that rises with
depth before eventually declining. Here we construct a profile representative
of such a lake: ~5% TOC at the surface, rising to ~15% in the mid-Holocene,
and returning to ~5% by the early Holocene.

```{r lake_toc_profile, fig.height = 4}
# Age-depth model: 50 cm/ka, 10 ka core
# Surface (0 ka) = 0 cm; mid-Holocene (5 ka) = 250 cm; base (10 ka) = 500 cm
am_lake <- rate_to_age_model(rate_cm_per_ka = 50, max_depth_cm = 500)

depths_lake <- c(  0,  50, 150, 250, 350, 450, 500)
toc_lake    <- c(  5,   7,  12,  15,  10,   6,   5)
toc_lake_fn <- make_toc_model(depth_cm = depths_lake, toc_pct = toc_lake)

plot_depths <- seq(0, 500, by = 5)
data.frame(
  depth_cm = plot_depths,
  age_ka   = am_lake$depth_to_age(plot_depths),
  toc      = toc_lake_fn(plot_depths)
) |>
  ggplot(aes(x = toc, y = depth_cm)) +
  geom_line(colour = "sienna", linewidth = 1) +
  geom_point(data = data.frame(depth_cm = depths_lake, toc = toc_lake),
             colour = "sienna", size = 2) +
  scale_y_reverse(
    sec.axis = sec_axis(~ . / 50, name = "Age (ka BP)")
  ) +
  labs(x = "TOC (%)", y = "Depth (cm)",
       title = "Lake sediment TOC profile: mid-Holocene productivity peak") +
  theme_bw()
```

Notice that TOC at 250 cm (mid-Holocene) is three times the surface value.
Because the model normalises bacterial resetting to surface TOC, the
`TOC_z / TOC_0` ratio reaches 3 at that depth, making bacterial resetting
*three times stronger* there than at the surface.

### D/L profiles under different resetting rates

```{r lake_dl_profiles}
depths <- seq(0, 500, by = 10)

f_bac_vals <- c(0, 1e-4, 1e-3, 5e-3)
labels     <- c("f_bac = 0 (abiotic)", "f_bac = 1e-4", "f_bac = 1e-3", "f_bac = 5e-3")

res_lake <- lapply(seq_along(f_bac_vals), function(j) {
  out <- racemize_depth_series(depths,
                               age_model  = am_lake,
                               temp_model = 5,
                               toc_model  = toc_lake_fn,
                               f_bac      = f_bac_vals[j])
  out$scenario <- labels[j]
  out
})

do.call(rbind, res_lake) |>
  mutate(scenario = factor(scenario, levels = labels)) |>
  ggplot(aes(x = DL, y = depth_cm, colour = scenario)) +
  geom_line(linewidth = 1) +
  scale_y_reverse() +
  scale_colour_brewer(palette = "Reds") +
  labs(x = "D/L", y = "Depth (cm)", colour = NULL,
       title = "Humped TOC creates a D/L dip at mid-Holocene depths",
       subtitle = "T = 5 deg C constant; strongest resetting at ~250 cm (5 ka)") +
  theme_bw() +
  theme(legend.position = "bottom")
```

The D/L profiles develop a visible inflection or dip near 250 cm at higher
`f_bac` values. The curve is no longer monotonically increasing with depth.
At 500 cm the profiles converge back toward the abiotic case because TOC has
returned to its surface value (ratio = 1) and the early-Holocene sediment
experienced lower resetting rates throughout its burial history.

### The spurious temperature signal

This is the key problem for paleothermometry: an abiotic inversion applied
to a bacterially reworked D/L profile would interpret the mid-Holocene D/L
dip as a real temperature signal. We can estimate the implied apparent
temperature at each depth using the same grid-search approach as above.

```{r lake_temp_bias}
T_grid <- seq(1, 10, by = 0.1)

compute_apparent_T <- function(toc_fn, f_bac_val, true_T = 5) {
  bac_out <- racemize_depth_series(depths, am_lake,
                                   temp_model = true_T,
                                   toc_model  = toc_fn,
                                   f_bac      = f_bac_val)
  apparent <- sapply(seq_along(depths), function(di) {
    target <- bac_out$DL[di]
    if (target == 0) return(NA_real_)
    ab_DL <- sapply(T_grid, function(T) {
      racemize_depth_series(depths[di], am_lake, temp_model = T)$DL
    })
    T_grid[which.min((ab_DL - target)^2)]
  })
  data.frame(depth_cm   = depths,
             age_ka     = am_lake$depth_to_age(depths),
             apparent_T = apparent,
             bias_C     = apparent - true_T,
             toc        = toc_fn(depths))
}

bias_lake <- bind_rows(
  mutate(compute_apparent_T(toc_lake_fn, 1e-3), scenario = "f_bac = 1e-3"),
  mutate(compute_apparent_T(toc_lake_fn, 5e-3), scenario = "f_bac = 5e-3")
)

ggplot(bias_lake |> filter(!is.na(apparent_T)),
       aes(x = bias_C, y = depth_cm, colour = scenario)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 1) +
  scale_y_reverse() +
  scale_colour_manual(values = c("steelblue", "tomato")) +
  labs(x = "Apparent temperature bias (deg C)", y = "Depth (cm)",
       colour = NULL,
       title = "Non-monotonic TOC creates spurious mid-Holocene cooling",
       subtitle = "True temperature is constant 5 deg C throughout") +
  theme_bw() +
  theme(legend.position = "bottom")
```

The apparent temperature is not simply cooler everywhere: it has a structured
depth pattern that mirrors the TOC hump. Near the surface, apparent
temperature is slightly cool (resetting rate proportional to 5% TOC). In the
mid-Holocene, bacterial resetting is strongest (15% TOC, ratio = 3), producing
the largest apparent cooling, up to several degrees C for `f_bac = 5e-3`.
Below 250 cm, as TOC declines back to 5%, the bias shrinks.

The practical implication: **a Holocene productivity event can masquerade as
a Holocene temperature event in the D/L record**. If a lake has independent
evidence of mid-Holocene high productivity (e.g., high biogenic silica, high
organic matter flux), this correction should be applied before attempting a
temperature reconstruction.

## Practical guidance

- **If TOC data are available**, construct a `toc_model` with `make_toc_model()`
  and run sensitivity analyses across `f_bac = c(0, 1e-4, 1e-3, 5e-3)` to
  bracket the plausible range of bacterial influence.

- **Marine sediments** (Braun et al. 2017 context): bacterial effects are
  significant, with `f_bac` likely 1e-3 to 1e-2 yr^-1.

- **Cold, oligotrophic lake sediments** (low TOC, low temperature):
  bacterial activity is much lower. TOC < 1% and temperatures near 4 deg C
  suggest `f_bac` closer to 1e-4 yr^-1 or smaller.

- **Productive lake sediments with non-monotonic TOC**: as shown above, a
  mid-Holocene productivity peak can create a structured apparent temperature
  signal that resembles real Holocene climate variability. If independent
  proxies (biogenic silica, organic matter flux, pollen) indicate past
  productivity changes, a measured downcore TOC profile should be used rather
  than a simple decay model.

- **Default behaviour** (`f_bac = 0`): pure abiotic racemization, identical to
  all prior results. Bacterial correction is strictly opt-in.

- Braun et al. (2017) note that D/L ratios in marine sediments are always
  lower than pure racemization predicts. In Holocene lake sediments where
  D/L ratios have had less time to accumulate, this bias may be relatively
  small compared to the temperature signal of interest, but should be checked.
```
