---
title: "Bayesian Kinetics Estimation (Stage 1)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Kinetics Estimation (Stage 1)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
library(AARP)
library(ggplot2)
library(dplyr)
```

## Overview

Stage 1 estimates the three AAR kinetic parameters ($A_e$: pre-exponential
factor; $E_a$: activation energy; $x$: power-law exponent) from two
data sources:

1. **Lab heating experiments**: surface sediment heated at 80, 100, and 110°C
   for known durations. Temperature is controlled; time is controlled; only
   the resulting D/L is measured. These data tightly constrain the Arrhenius
   relationship.

2. **Downcore data from 4°C lakes**: lakes whose hypolimnion stays at a
   near-constant 4°C year-round (deep, cold-climate lakes). Because
   temperature is effectively known, the downcore D/L profile constrains how
   D/L accumulates through time (i.e., the power-law exponent $x$).

Together, these two data types break the degeneracy among the three parameters.

We work in transformed space: $\log(A_e)$ (since $A_e \approx e^{42}$ is
numerically unwieldy) and $\log(\sigma)$ for the observation noise (enforces
positivity without hard truncation).

## Generate synthetic data

For this prototype we use synthetic data with known true parameters so we can
verify recovery.

```{r generate_data}
set.seed(42)
true_AAR <- default_AAR_params  # Ae=exp(42.12), Ea=31.5, x=3

# Heating experiments: 3 temps × 8 times × 3 replicates = 72 observations
heating_data <- sim_heating_data(AAR_params = true_AAR, sigma = 0.01, seed = 1)

# Calibration core: 30 samples at 500-yr spacing to 15 ka in a 4°C lake
am_cal   <- rate_to_age_model(rate_cm_per_ka = 50, max_depth_cm = 750)
cal_data <- sim_downcore_data(am_cal,
                              temp_model     = 4,
                              sample_ages_ka = seq(0.5, 15, by = 0.5),
                              AAR_params     = true_AAR,
                              sigma          = 0.015,
                              seed           = 2)
```

### Visualise the synthetic observations

```{r plot_obs, fig.height=4}
# Heating experiments
ggplot(heating_data, aes(x = time_yr, y = DL_obs,
                          color = factor(temp_C), group = factor(temp_C))) +
  geom_point(size = 1.5) +
  geom_line(aes(y = DL_pred), linewidth = 0.8) +
  scale_x_log10() +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Heating duration (yr, log scale)", y = "D/L",
       color = "Temp (°C)", title = "Synthetic heating experiment data") +
  theme_bw()
```

```{r plot_cal, fig.height=3.5}
# Calibration core
ggplot(cal_data, aes(x = DL_obs, y = depth_cm)) +
  geom_point(size = 1.5, colour = "steelblue") +
  geom_line(aes(x = DL_pred), colour = "grey50") +
  scale_y_reverse() +
  labs(x = "D/L", y = "Depth (cm)",
       title = "Synthetic 4°C calibration core") +
  theme_bw()
```

## Prior specification

The log-posterior combines two likelihoods (heating experiments and the
calibration core) with a **prior** on the four sampled parameters. Inspecting
the priors before running the sampler confirms they are broad relative to the
data so the observations drive the posterior.

| Parameter | Prior | Centre | SD |
|-----------|-------|--------|----|
| $\log(A_e)$ | Normal | 42.12 | 2 |
| $E_a$ (kcal/mol) | Normal | 31.5 | 5 |
| $x$ | Normal | 3 | 0.5 |
| $\log(\sigma)$ | Normal | $\log(0.02)$ | 1 |

Centres are drawn from published Ile-Ile estimates (Kaufman 2000). Standard
deviations are wide enough to allow the data to move the posterior substantially,
while still ruling out physically implausible values.

```{r prior_kin, fig.height=5}
prior_pars <- list(
  log_Ae    = c(mean = 42.12,     sd = 2),
  Ea        = c(mean = 31.5,      sd = 5),
  x         = c(mean = 3,         sd = 0.5),
  log_sigma = c(mean = log(0.02), sd = 1)
)

prior_df <- do.call(rbind, lapply(names(prior_pars), function(nm) {
  p <- prior_pars[[nm]]
  v <- seq(p["mean"] - 3.5 * p["sd"], p["mean"] + 3.5 * p["sd"],
           length.out = 400)
  data.frame(parameter = nm, value = v,
             density   = dnorm(v, p["mean"], p["sd"]))
}))

ggplot(prior_df, aes(x = value, y = density)) +
  geom_area(fill = "darkorange", alpha = 0.3) +
  geom_line(colour = "darkorange", linewidth = 0.9) +
  facet_wrap(~parameter, scales = "free", ncol = 2) +
  labs(x = "Value", y = "Prior density",
       title = "Stage 1 priors: kinetic parameters") +
  theme_bw()
```

## Run Stage 1 MCMC

We run a random-walk Metropolis-Hastings chain. Proposal standard deviations
are set to target roughly 25% acceptance rate. Check `acceptance_rate` in
the output and adjust `prop_kin` if needed.

```{r mcmc_stage1, cache=TRUE}
init_kin <- c(log_Ae    = 42.12,
              Ea        = 31.5,
              x         = 3.0,
              log_sigma = log(0.01))

prop_kin <- c(log_Ae    = 0.005,
              Ea        = 0.05,
              x         = 0.003,
              log_sigma = 0.008) / 2

mcmc_kin <- run_mcmc(log_posterior_kinetics,
                     init        = init_kin,
                     n_iter      = 10000,
                     proposal_sd = prop_kin,
                     heating_data      = heating_data,
                     downcore_cal_data = cal_data,
                     age_model_cal     = am_cal,
                     temp_cal          = 4)

cat("Acceptance rate:", round(mcmc_kin$acceptance_rate * 100, 1), "%\n")
```

## Trace plots

Burn-in (grey) should be discarded before summarising the posterior. Chains
should look like white noise after burn-in, with no trend or drift.

```{r traces, fig.height=7}
plot_mcmc_chains(mcmc_kin, burnin = 1000)
```

## Posterior distributions

Red dashed lines show the true parameter values used to generate the data.
Orange curves show the prior density. The posterior (blue fill) should be much
narrower than the prior, confirming that the data are informative, and centred
near the true values.

```{r posteriors}
true_kin <- c(log_Ae    = 42.12,
              Ea        = 31.5,
              x         = 3.0,
              log_sigma = log(0.01))

plot_posterior_kinetics(mcmc_kin, true_params = true_kin, burnin = 1000) +
  geom_line(data = prior_df, aes(x = value, y = density),
            colour = "darkorange", linewidth = 0.9, inherit.aes = FALSE)
```

## Posterior predictive check

Draw 200 parameter sets from the posterior and run the forward model to
confirm the fitted parameters reproduce the calibration core observations:

```{r ppc}
depths_check <- seq(0, 750, by = 15)
posterior_predictive_DL(mcmc_kin,
                        age_model     = am_cal,
                        depths_cm     = depths_check,
                        temp_model    = 4,
                        observed_data = cal_data,
                        burnin        = 1000,
                        n_draws       = 200)
```

Points are the synthetic observations; the shaded ribbon is the 95% posterior
predictive interval. Good coverage of the points confirms the model fits the
data. These kinetics estimates are passed to Stage 2.
```
