---
title: "Holocene Temperature Reconstruction (Stage 2)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Holocene Temperature Reconstruction (Stage 2)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
library(AARP)
library(ggplot2)
library(dplyr)
```

## Overview

Stage 2 takes the kinetic parameters estimated in Stage 1 as fixed and infers
the temperature history of a new ("target") core whose thermal history is
unknown. The unknown is parameterised as a vector of temperatures at 6 control
knots (0, 2, 4, 6, 8, 10 ka BP) connected by linear interpolation. A
first-difference smoothness prior penalises large step-to-step changes.

This two-stage approach avoids simultaneously inferring kinetics and
temperature from the same data, which would be highly under-determined.

## Stage 1 recap: retrieve kinetics

We regenerate the Stage 1 MCMC with a fixed seed so the kinetics are
reproducible across vignettes:

```{r stage1, cache=TRUE}
true_AAR <- default_AAR_params

# Heating experiments
heating_data <- sim_heating_data(AAR_params = true_AAR, sigma = 0.01, seed = 1)

# Calibration core (4°C lake)
am_cal   <- rate_to_age_model(rate_cm_per_ka = 50, max_depth_cm = 750)
cal_data <- sim_downcore_data(am_cal,
                              temp_model     = 4,
                              sample_ages_ka = seq(0.5, 15, by = 0.5),
                              AAR_params     = true_AAR,
                              sigma          = 0.015,
                              seed           = 2)

init_kin <- c(log_Ae = 42.12, Ea = 31.5, x = 3.0, log_sigma = log(0.01))
prop_kin <- c(log_Ae = 0.05,  Ea = 0.3,  x = 0.05, log_sigma = 0.1) / 5

mcmc_kin <- run_mcmc(log_posterior_kinetics,
                     init              = init_kin,
                     n_iter            = 10000,
                     proposal_sd       = prop_kin,
                     heating_data      = heating_data,
                     downcore_cal_data = cal_data,
                     age_model_cal     = am_cal,
                     temp_cal          = 4)

cat("Stage 1 acceptance rate:", round(mcmc_kin$acceptance_rate * 100, 1), "%\n")
```

Extract the posterior median for the kinetic parameters to use as fixed values
in Stage 2:

```{r kinetics_extract}
burnin_idx   <- 1001:10000
kin_post     <- mcmc_kin$samples[burnin_idx, ]

kin_fixed <- c(
  log_Ae    = median(kin_post[, "log_Ae"]),
  Ea        = median(kin_post[, "Ea"]),
  x         = median(kin_post[, "x"]),
  log_sigma = median(kin_post[, "log_sigma"])
)

cat("Fixed kinetics used in Stage 2:\n")
print(round(kin_fixed, 4))
```

## Synthesise the target core

The target core comes from a lake with an unknown temperature history. For
this demonstration we use a linear Holocene cooling scenario: 8°C at 10 ka BP
declining to 5°C at present. Samples span 0.5 to 10 ka at 500-yr spacing.

```{r target_core}
true_T_history <- data.frame(age_ka = c(0, 10), temp_C = c(5, 8))

am_target    <- rate_to_age_model(rate_cm_per_ka = 40, max_depth_cm = 400)
target_data  <- sim_downcore_data(am_target,
                                   temp_model     = true_T_history,
                                   sample_ages_ka = seq(0.5, 10, by = 0.5),
                                   AAR_params     = true_AAR,
                                   sigma          = 0.015,
                                   seed           = 3)

ggplot(target_data, aes(x = DL_obs, y = depth_cm)) +
  geom_point(size = 1.5, colour = "steelblue") +
  geom_line(aes(x = DL_pred), colour = "grey50") +
  scale_y_reverse() +
  labs(x = "D/L", y = "Depth (cm)",
       title = "Synthetic target core (true cooling: 8°C → 5°C)") +
  theme_bw()
```

## Prior specification for Stage 2

```{r knot_setup}
knot_ages <- c(0, 2, 4, 6, 8, 10)  # ka BP
```

Stage 2 has three prior components for the temperature history parameters:

**Marginal temperature prior** — each knot temperature receives a weakly
informative Normal(5, 3) prior, centred near present-day Arctic/sub-Arctic
mean annual temperatures. The ±3σ range spans roughly −4°C to 14°C, which
easily encompasses Holocene variability without constraining the inference
to any particular trend.

**Smoothness prior** — adjacent knot differences receive a Normal(0, 2)
penalty on the log scale. This encodes the physical expectation of gradual
Holocene temperature change: 2-ka step sizes larger than ~4°C (2σ) are
strongly penalised but not impossible.

**Noise prior** — log(σ) receives Normal(log(0.02), 1), the same as Stage 1,
centred near HPLC analytical precision.

```{r prior_stage2, fig.height=3.5}
# Marginal temperature prior and smoothness prior
prior_T_df <- rbind(
  data.frame(
    component     = "Knot temperature:  N(5, 3)",
    value         = seq(-4, 14, length.out = 300),
    density       = dnorm(seq(-4, 14, length.out = 300), 5, 3)
  ),
  data.frame(
    component     = "Step size (2-ka):  N(0, 2)",
    value         = seq(-7, 7, length.out = 300),
    density       = dnorm(seq(-7, 7, length.out = 300), 0, 2)
  )
)

ggplot(prior_T_df, aes(x = value, y = density)) +
  geom_area(fill = "darkorange", alpha = 0.3) +
  geom_line(colour = "darkorange", linewidth = 0.9) +
  facet_wrap(~component, scales = "free", ncol = 2) +
  labs(x = "Value (°C)", y = "Prior density",
       title = "Stage 2 priors: temperature history") +
  theme_bw()
```

Drawing 200 independent trajectories from the marginal knot prior gives a
prior predictive envelope — the range of temperature histories considered
plausible before the D/L data are observed:

```{r prior_predictive_T, fig.height=4}
set.seed(7)
n_prior     <- 200
prior_T_mat <- matrix(rnorm(n_prior * length(knot_ages), mean = 5, sd = 3),
                      nrow = n_prior)

summ_prior <- do.call(rbind, lapply(seq_along(knot_ages), function(j) {
  data.frame(
    age_ka = knot_ages[j],
    lo95   = quantile(prior_T_mat[, j], 0.025),
    hi95   = quantile(prior_T_mat[, j], 0.975),
    lo50   = quantile(prior_T_mat[, j], 0.25),
    hi50   = quantile(prior_T_mat[, j], 0.75)
  )
}))

ggplot(summ_prior, aes(x = age_ka)) +
  geom_ribbon(aes(ymin = lo95, ymax = hi95),
              fill = "darkorange", alpha = 0.2) +
  geom_ribbon(aes(ymin = lo50, ymax = hi50),
              fill = "darkorange", alpha = 0.4) +
  scale_x_reverse() +
  labs(x = "Age (ka BP)", y = "Temperature (°C)",
       title = "Prior predictive envelope: temperature history",
       subtitle = "Inner band: 50% prior interval; outer band: 95% prior interval") +
  theme_bw()
```

## Run Stage 2 MCMC

The temperature parameterisation uses 6 control knots. Initialise near a
plausible flat temperature and use a moderately diffuse proposal:

```{r stage2, cache=TRUE}
init_T <- c(T_0ka  = 6, T_2ka  = 6, T_4ka  = 6,
            T_6ka  = 6, T_8ka  = 6, T_10ka = 6,
            log_sigma = log(0.015))

prop_T <- c(T_0ka  = 0.3, T_2ka  = 0.3, T_4ka  = 0.3,
            T_6ka  = 0.3, T_8ka  = 0.3, T_10ka = 0.3,
            log_sigma = 0.1) * 2

mcmc_T <- run_mcmc(log_posterior_temperature,
                   init              = init_T,
                   n_iter            = 10000,
                   proposal_sd       = prop_T,
                   knot_ages_ka      = knot_ages,
                   kinetics_params   = kin_fixed,
                   downcore_target_data = target_data,
                   age_model_target  = am_target)

cat("Stage 2 acceptance rate:", round(mcmc_T$acceptance_rate * 100, 1), "%\n")
```

## Trace plots

Check that chains mix well across all temperature knots:

```{r traces, fig.height=9}
plot_mcmc_chains(mcmc_T, burnin = 1000)
```

## Temperature reconstruction

The reconstruction shows the posterior median (line), the 95% credible
interval (blue ribbon), the true temperature history (red dashed line), and
the 95% prior predictive interval (orange ribbon). The posterior should be
substantially narrower than the prior, demonstrating that the D/L data are
informative about temperature:

```{r reconstruction, fig.height=4.5}
true_T_df <- data.frame(age_ka = knot_ages,
                        temp_C = c(5, 5.6, 6.2, 6.8, 7.4, 8))

plot_temperature_reconstruction(mcmc_T,
                                knot_ages_ka   = knot_ages,
                                true_T_history = true_T_df,
                                burnin         = 1000) +
  geom_ribbon(data = summ_prior,
              aes(x = age_ka, ymin = lo95, ymax = hi95),
              fill = "darkorange", alpha = 0.15, inherit.aes = FALSE) +
  labs(subtitle = "Orange: 95% prior; blue: 95% posterior")
```

The true cooling trend (red dashed line) should fall within the credible
envelope. Uncertainty is largest at the oldest end of the record, where the
cumulative integration of noisy D/L data limits resolution. The contrast
between the wide orange prior and the narrow blue posterior quantifies how
much information the D/L profile contributes.

## Posterior predictive check

Verify that the recovered temperature history reproduces the observed D/L
profile. For Stage 2 we draw temperature knots from the posterior, build a
temperature history per draw, and run the forward model with the fixed
kinetics:

```{r ppc}
post_T  <- tail(mcmc_T$samples, nrow(mcmc_T$samples) - 1000)
T_cols  <- grep("^T_", colnames(post_T), value = TRUE)
idx     <- sample(nrow(post_T), 200)

kin_AAR <- list(Ae = exp(kin_fixed[["log_Ae"]]),
                Ea = kin_fixed[["Ea"]],
                R  = 0.001987,
                x  = kin_fixed[["x"]])

depths_check <- seq(0, 400, by = 10)

pred_list <- lapply(idx, function(i) {
  tmp_model <- data.frame(age_ka = knot_ages,
                          temp_C = as.numeric(post_T[i, T_cols]))
  fwd       <- racemize_depth_series(depths_check, am_target, tmp_model,
                                     AAR_params = kin_AAR)
  fwd$draw  <- i
  fwd
})
pred_df <- do.call(rbind, pred_list)

summ <- pred_df |>
  dplyr::group_by(depth_cm) |>
  dplyr::summarise(median = median(DL),
                   lo95   = quantile(DL, 0.025),
                   hi95   = quantile(DL, 0.975),
                   .groups = "drop")

ggplot(summ, aes(x = median, y = depth_cm)) +
  geom_ribbon(aes(xmin = lo95, xmax = hi95, y = depth_cm),
              fill = "steelblue", alpha = 0.3) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(data = target_data, aes(x = DL_obs, y = depth_cm),
             colour = "black", size = 1.5) +
  scale_y_reverse() +
  labs(x = "D/L", y = "Depth (cm)",
       title = "Posterior predictive D/L profile") +
  theme_bw()
```

## Identifiability and uncertainty

Several factors govern the width of the credible interval:

- **Number of samples**: more observations reduce uncertainty. Here we used 20 samples.

- **Observation noise** (`sigma`): analytical precision in the D/L measurement. Here we used $\sigma \approx 0.01$ to $0.02$, is that realistic?

- **Kinetics uncertainty**: I fixed kinetics at their Stage 1 posterior median. Propagating the full Stage 1 posterior (i.e., drawing both kinetics and temperature jointly) would widen the credible interval further. We should do this, and turn this into a proper Bayesian Hierarchical Model.

```
